Kratos: blockchain solution for data privacy, literacy, and student agency in a data-driven educational ecosystem

### 2019-06-03 paper.tex
add images and changes

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
 \geometry{
 top=30mm,
 left=1.5in,
 right=1.5in
 }
\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}

@@

To improve work, school practitioners at local, district, and federal levels need data interoperability and educational information from data mining [Gates Foundation, 2015; US Dept of Education, 2012; ACT Policy Report, 2015]. This need originally emanates from the United States federal law which mandates schools at district, state, and federal level to collect student data to serve as 'accountability' metric for assessing school effectiveness [Elementary and Secondary Education Act, 2001; 2010].

@@

Interoperability challenges between vendors and schools at local, district, or federal level, further pose barriers not only to cohesive data management and sharing but also lacks technological privacy infrastructure and accountability [Zeide, 2014; Common Sense Media Research, 2018]. While digitizing and collecting student data is not new [Fitzgerald, 2014], technological developments like cloud computing and the growing use of web applications, learning management systems that gather fine-grained student data, amplify the concerns about data transfer, storage, use, and analysis [Sultan, 2010].

@@

Increasingly digitized learning environments enable constant data collection and algorithm-based assessment which change core school functions of teaching, assessment, and credentialing [Zeide, 2017]. Digital systems for teaching and assessment drive data-based decision-making and enable micromanagement of students that can further pose restrictions on teacher autonomy and student and parent participation and willingness to challenge education decision-making. Interactive learning environments allow students to manage coursework online, take tests, and use a variety of online learning material. Schools deploy management systems for sorting and storing reporting and operational data. While some data is stored on local servers at school level, other school data is managed by third-party vendors [Zhang, Li, and Hao, 2015], who may consider data proprietary and thus inaccessible. This puts additional barriers to effective data sharing and comprehensive understanding of school data.

@@

The present challenges of fragmented data access and use, lack of student agency, accountability and auditability to data access and use demand a socio-technological solution, one which can ensure that educational data can be used to help improve school processes while student privacy, agency, and future opportunities are not diminished.

@@

In this work we propose the development of a blockchain structure for student data management, privacy, accountability, and auditability. We optimise on available data standards and construct references to disparate data across different systems and applications on a decentralised ledger. Our system organises these references onto an integral structure of student data and complements the existing educational data standards and interoperability in three distinctive ways. Through on-the-network permissioning, we enable school and student agency with data accuracy, auditability, and sharing. We design modular APIs to integrate with existing school systems and data standards. Lastly, we build a simple user interface that gives students/parents and schools access control of records enabling control over what portion or part of the data can be shared. Additionally, the interface enables applied data literacy and an application with social functionality for student feedback.

@@

We present Kratos not as the only solution to lack of data interoperability, accountability, and privacy but rather as a system that complements existing efforts and solutions developed by the various stakeholders in the educational ecosystem (Data Quality Standards; Project Unicorn).

@@

\section{Existing problems}
In order to have a better picture on the problems schools currently face, Kratos has partnered with the Cambridge Public School (CPS) District, the controlling body for all schools in Cambridge, Massachusetts. With the help of the Chief Information Officer, Chief Technical Officer and Database Administrator, we explored problems and ways to solve them along with feedback on potential mitigation schemes:

\begin{itemize}
  \item Data access - CPS has agreements with over 150 partners ([7]) and on account, none of these providers provide direct access to data. Data access is instead provided in the form of a report or a summary which the district database administrator can request and download. Some providers are even less flexible and only offer to give reports once in 45 days making it impossible for teachers to act on feedback that they might infer from the data.
  \item Lack of standards - The different vendors that CPS partners have their own implementation, most of them not regualted by any standards. The schools themselves follow a common standard defined by the SIF but there is no way to enforce this upon vendors. Even if the vendor agreed to follow a standard defined by the SIF, it is not possible to concretely verify this claim.
  \item Lack of transparency - The vendors do not provide a comprehensive list of the data they collect from students. This makes data audits extremely difficult since we don't exhaustively know what to audit.
  \item Security - It is impossible to verify if vendors are following best practices in data storage and retrieval. This is extremely dangerous since a malicious entity can arbitrarily change or retrieve student records from a poorly maintained database
\end{itemize}

\section{Kratos}
Kratos aims to define a set of principles and guidelines that schools should follow and provides a platform for both schools and students to see what kind of data is being collected by vendors along with ways to tweak data access. The goal for students would be to see how their data is being used and above the age of 18, gain acccess to this data. We define this as Data Ownership and Visibility- how students can own the data after the age of 18 and how students can see what their data is being used for. The goal for the school would be to make disparate data from different vendors compatible with each other and we define this problem as Data Interoperability.

\subsection{Data ownership}
In conventional systems, data ownership can be proved with the help of an access token but this provides no guarantee on when the owner came into possession of the data. In order to attest ownership at a specific point in time, we need timestamping services like those described in \hyperref[sec:1]{[1]}, \hyperref[sec:2]{[2]} and \hyperref[sec:3]{[3]}. A timestamping service requires something to be committed along with the timestamp and Kratos envisions this to be a cryptographic hash which also acts as an access token which vendors can use to access the data.
\bigbreak
The aim of using a cryptographic hash like SHA-3 \hyperref[sec:6]{[6]} is to ensure a uniquely random reference to the data to avoid out of channel data leaks. Kratos suggests using an element of randomness like a salt to generate the hash in order to have the ability to revoke tokens by regenerating randomness. Kratos enforces that all schools encrypt their data before creating access tokens to mitigate the risk of loss/theft of data. Past studies like \hyperref[sec:4]{[4]} and \hyperref[sec:5]{[5]} show that firms are willing to circumnavigate laws to collect data and encrypting data by design ensures that no third party can have access without being granted so explicitly.
\bigbreak
Kratos does not specify at which level data needs to be encrypted - it can be at the school level or at the student level depending on existing frameworks and rules surrounding the school. If the school chooses to encrypt student data on behalf of the student, Kratos enforces that the school use a unique key for each individual to minimize the risk of key theft. In addition to this, Kratos suggests that schools store their encrypted data in a distributed file storage system like IPFS to ensure data redundancy and availability in case of a mishap. Storing data on IPFS also makes it easier to create timestamps since it is sufficient to reference the IPFS pointer instead of potentially hashing the whole data.
\bigbreak
In the event a user wants to revoke access to a particular vendor, he could do so by changing the encryption key or by changing the randomness used to generate the access token. We suggest users don't regenerate encryption keys but Kratos does not enforce this and will provide end users with the option to choose between the two.
\bigbreak
Since schools have their own sets of policies, Kratos doesn't strictly enforce a set of practices for users to follow. This ensures that adoption of Kratos isn't constrained by a certain set of rules. At the same time, Kratos defines a set of minimum requirements to be on board to ensure good practices on date protection are followed.
\bigbreak
Kratos also doesn't specify how and where the commitments need to be stored and leaves it to schools to provide their feedback. Potential solutions include a centralized timestamping server, a permissioned blockchain with the different schools as the nodes and simple timestamping commitments to an existing blockchain. All three options have their benefits and constraints and Kratos would arrive at a final recommendation depending on the option the majority of schools are convenient with
\bigbreak

@@

\bigbreak
Kratos proposes that this be done in the form of a templated script that can be written for every vendor. Building a common script for all the vendors poses a challenge since the number of vendors and the different fields and standards they follow are ever changing. This templated script would be trigerred automatically by Kratos each time it receives an incoming report and the report itself will be parsed ot udnerstand which vendor it originates from.

@@

\section{References}
\begin{enumerate}
    \label{sec:1}
    \item Todd, P. OpenTimestamps: Scalable, Trustless, Distributed Timestamping with Bitcoin (2016). URL \url{https://petertodd.org/2016/opentimestamps-announcement}
    \label{sec:2}
    \item Szalachowski, P. (2018). Towards more reliable Bitcoin timestamps. arXiv preprint arXiv:1803.09028.
    \label{sec:3}
    \item Massias, H., Avila, X. S., \& Quisquater, J. J. (1999). Design of a secure timestamping service with minimal trust requirement. In the 20th Symposium on Information Theory in the Benelux.
    \label{sec:4}
    \item Zetter, K. Google Collected Data on Schoolchildren without permission (2016). URL \url{https://www.wired.com/2015/12/google-collected-data-on-schoolchildren-without-permission/}
    \label{sec:5}
    \item Dissent, Back-To-School Revolt in Springfield? Employees balk over using Google Drive as evidence of massive privacy breach mounts (2018).  URL \url{https://www.pogowasright.org/back-to-school-revolt-in-springfield-employees-balk-over-using-google-drive-as-evidence-of-massive-privacy-breach-mounts/}
    \label{sec:6}
    \item Bertoni, G., Daemen, J., Peeters, M., Van Assche, G., \& Van Keer, R. (2012). Keccak implementation overview. URL  \url{http://keccak.neokeon.org/Keccak-implementation-3.2.pdf}.
    \label{sec:7}
    \item CPSD. District Agreements Listing (2019). URL  \url{https://sdpc.a4l.org/district_listing.php?districtID=457}
    \label{sec:8}
    \item Colins, A., and Halverson, R. (2018).\textit{ Re-thinking education in the age of technology: The digital revolution and schooling in America}. Teach Zeide, E. (2016).
    \label{sec:9}
    \item 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?.
    \label{sec:10}
    \item Colins, A., and Halverson, R. (2018).\textit{ Re-thinking education in the age of technology: The digital revolution and schooling in America}. Teachers College Press: New York.
    \label{sec:11}
    \item Cody, A. (2013). Will the data warehouse become every student and teacher's 'permanent record'? \textit{Education Week}, May, 20.
    \label{sec:12}
    \item Gibson, D. C., Webb, M., and Ifenthaler, D. (2015). Challenges of big data in educational assessment.\textit{Proceedings of the IADIS International Conference of Exploratory Learning in Digital Age}, 92-100.
    \label{sec:13}
    \item Sultan, N. (2010). Cloud computing for education: a new dawn? 30 International Journal of Information Management  109.
    \label{sec:14}
    \item Fitzgerald, B. (2014). Data collection isn't new. And it predates common core.Funny Monkey.
    \label{sec:15}
    \item Zeide, E. (2016). 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?.
\end{enumerate}
\end{document}

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\usepackage[final]{graphicx}
\usepackage{grffile}
 \geometry{
 top=30mm,
 left=1.5in,
 right=1.5in
 }
\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}

@@

To improve work, school practitioners at local, district, state, and federal levels need data interoperability and educational information from data mining [Gates Foundation, 2015; US Dept of Education, 2012; ACT Policy Report, 2015]. This need originally emanates from the United States federal law which mandates schools at district, state, and federal level to collect student data to serve as 'accountability' metric for assessing school effectiveness [Elementary and Secondary Education Act, 2001; 2010].

@@

Interoperability challenges between vendors and schools at local, district, state, or federal level further pose barriers not only to cohesive data management and sharing but also lacks technological privacy infrastructure and accountability [Zeide, 2014; Common Sense Media Research, 2018]. While digitizing and collecting student data is not new [Fitzgerald, 2014], technological developments like cloud computing and the growing use of web applications, learning management systems that gather fine-grained student data, amplify the concerns about data transfer, storage, use, and analysis [Sultan, 2010].

@@

Increasingly digitized learning environments enable constant data collection and algorithm-based assessment which change core school functions of teaching, assessment, and accreditation [Zeide, 2017]. Digital systems for teaching and assessment drive data-based decision-making and enable micromanagement of students that can further pose restrictions on teacher autonomy and student (and parent) participation and willingness to challenge education decision-making. Interactive learning environments allow students to manage coursework online, take tests, and use a variety of online learning material. Schools deploy management systems for sorting and storing reporting and operational data. While some data is stored on local servers at school district level, other school data is managed by third-party vendors [Zhang, Li and Hao, 2015], who may consider data proprietary and thus inaccessible. This puts additional barriers to effective data sharing and comprehensive understanding of school data.

@@

The present challenges of fragmented data access and use, lack of student agency, accountability, and auditability to data access and use demand a socio-technological solution, one which can ensure that educational data can be used to help improve school processes while student privacy, agency, and future opportunities are not diminished.

@@

In this work we propose the development of a structure for student data management, privacy, accountability, and auditability. We build on top of existing data standards and construct a common data schema for disparate data across different systems. Our system organises these references onto an integral structure of student data and complements the existing educational data standards and interoperability in three distinctive ways. Through network permissioning and proofs of ownership on a distributed ledger, we enable data auditability and accountability. We design data analytic models to integrate with existing school systems and data standards. Lastly, we build a simple user interface that gives students, parents and schools access to disparate data, enabling control over what portion of the data can be shared. Additionally, we propose an application with social functionality for student feedback.

@@

We present Kratos not as an independent solution to lack of data interoperability, accountability, and privacy but rather as a system that complements existing efforts and solutions developed by the various stakeholders in the educational ecosystem (Data Quality Standards, 2018; Project Unicorn; Bill and Melinda Gates Foundation, 2015).

\section{Existing problems}
To contextualise the complexity related to school data - the opportunities and challenges schools face as a result of the growing digitisation of operation and academic processes - Kratos has partnered with  Cambridge Public School (CPS) district that administer public elementary and high schools in Cambridge, Massachusetts, the Access for Learning (SIF), and the Student Data Privacy Consortium. With the help of CPS's Information and Communication Technology Services Chief Information Officer and Database Administrator, we explored data interoperability issues with them along with feedback on potential mitigation schemes:

\begin{itemize}
  \item Data access - CPS is in agreement with over 100 vendors providing education technologies ([7]). The majority of these would mostly provide no direct and comprehensive access to data generated about and by students who make use of their products or services. Where available, data access is  provided in the form of reports - a summary of information which the district database administrator can request and download. Some providers are even less flexible and only offer to give reports once in 45 days making it impossible for teachers to act on feedback that they might infer from the data. On other occasions vendors supply data directly to teachers as well as the students in the form of digital dashboards. Where teachers obtain data it requires further work to convert it into meaningful information upon which the teacher can adjust and plan instruction (Data Quality Campaign, 2018). 
  \item Unified data standardization challenges - The different vendors that CPS partners have their own implementation, most of them not regulated by any standards. The schools themselves follow a common standard defined by the SIF but there is no way to enforce this upon vendors. Even if the vendor agreed to follow a standard defined by the SIF, it is not possible to concretely verify this claim.
  \item Lack of transparency - The vendors do not provide a comprehensive list of the data they collect from students. This makes data audits extremely difficult since we don't exhaustively know what to audit.
  \item Security - It is impossible to verify if vendors are following best practices in data storage and retrieval. This is extremely dangerous since a malicious entity can arbitrarily change or retrieve student records from a poorly maintained database
\end{itemize}

@@

\section{Kratos}
Kratos aims to define a set of principles and guidelines that schools should follow and provides a platform for both schools and students to see what kind of data is being collected by vendors along with ways to tweak data access. The goal for students would be to see how their data is being used and above the age of 18, gain access to this data. We define this as Data Ownership and Visibility- how students can own the data after the age of 18 and how students can see what their data is being used for. The goal for the school would be to make disparate data from different vendors compatible with each other and we define this problem as Data Interoperability.

\subsection{Data ownership}
In conventional systems, data ownership can be proved with the help of an access token but this provides no guarantee on when the owner came into possession of the data. In order to attest ownership at a specific point in time, we need time-stamping services like those described in \hyperref[sec:1]{[1]}, \hyperref[sec:2]{[2]} and \hyperref[sec:3]{[3]}. A time-stamping service requires something to be committed along with the timestamp and Kratos envisions this to be a cryptographic hash which also acts as an access token which vendors can use to access the data.
\bigbreak
The aim of using a cryptographic hash like SHA-3 \hyperref[sec:6]{[6]} is to ensure a uniquely random reference to the data to avoid out of channel data leaks. Kratos suggests using an element of randomness like a salt to generate the hash in order to have the ability to revoke tokens by regenerating randomness. Kratos enforces that all schools encrypt their data before creating access tokens to mitigate the risk of loss/theft of data. Past studies like \hyperref[sec:4]{[4]} and \hyperref[sec:5]{[5]} show that firms are willing to circumnavigate laws to collect data and encrypting data by design ensures that no third party can have access without being granted so explicitly.
\bigbreak
Kratos does not specify at which level data needs to be encrypted - it can be at the school level or at the student level depending on existing frameworks and rules surrounding the school. If the school chooses to encrypt student data on behalf of the student, Kratos enforces that the school use a unique key for each individual to minimize the risk of key theft. In addition to this, Kratos suggests that schools store their encrypted data in a distributed file storage system like IPFS to ensure data redundancy and availability in case of a mishap. Storing data on IPFS also makes it easier to create timestamps since it is sufficient to reference the IPFS pointer instead of potentially hashing the whole data.
\bigbreak
In the event a user wants to revoke access to a particular vendor, he could do so by changing the encryption key or by changing the randomness used to generate the access token. We suggest users don't regenerate encryption keys but Kratos does not enforce this and will provide end users with the option to choose between the two.
\bigbreak
Since schools have their own sets of policies, Kratos doesn't strictly enforce a set of practices for users to follow. This ensures that adoption of Kratos isn't constrained by a certain set of rules. At the same time, Kratos defines a set of minimum requirements to be on board to ensure good practices on date protection are followed.
\bigbreak
Kratos also doesn't specify how and where the commitments need to be stored and leaves it to schools to provide their feedback. Potential solutions include a centralized time-stamping server, a permissioned blockchain with the different schools as the nodes and simple time-stamping commitments to an existing blockchain. All three options have their benefits and constraints and Kratos would arrive at a final recommendation depending on the option the majority of schools are convenient with
\bigbreak

@@

\bigbreak
Kratos proposes that this be done in the form of a templated script that can be written for every vendor. Building a common script for all the vendors poses a challenge since the number of vendors and the different fields and standards they follow are ever changing. This templated script would be triggered automatically by Kratos each time it receives an incoming report and the report itself will be parsed to understand which vendor it originates from.

@@

\section{References}
\begin{enumerate}
    \label{sec:1}
    \item Todd, P. OpenTimestamps: Scalable, Trustless, Distributed Timestamping with Bitcoin (2016). URL \url{https://petertodd.org/2016/opentimestamps-announcement}
    \label{sec:2}
    \item Szalachowski, P. (2018). Towards more reliable Bitcoin timestamps. arXiv preprint arXiv:1803.09028.
    \label{sec:3}
    \item Massias, H., Avila, X. S., \& Quisquater, J. J. (1999). Design of a secure timestamping service with minimal trust requirement. In the 20th Symposium on Information Theory in the Benelux.
    \label{sec:4}
    \item Zetter, K. Google Collected Data on Schoolchildren without permission (2016). URL \url{https://www.wired.com/2015/12/google-collected-data-on-schoolchildren-without-permission/}
    \label{sec:5}
    \item Dissent, Back-To-School Revolt in Springfield? Employees balk over using Google Drive as evidence of massive privacy breach mounts (2018).  URL \url{https://www.pogowasright.org/back-to-school-revolt-in-springfield-employees-balk-over-using-google-drive-as-evidence-of-massive-privacy-breach-mounts/}
    \label{sec:6}
    \item Bertoni, G., Daemen, J., Peeters, M., Van Assche, G., \& Van Keer, R. (2012). Keccak implementation overview. URL  \url{http://keccak.neokeon.org/Keccak-implementation-3.2.pdf}.
    \label{sec:7}
    \item CPSD. District Agreements Listing (2019). URL  \url{https://sdpc.a4l.org/district_listing.php?districtID=457}
    \label{sec:8}
    \item Colins, A., and Halverson, R. (2018).\textit{ Re-thinking education in the age of technology: The digital revolution and schooling in America}. Teach Zeide, E. (2016).
    \label{sec:9}
    \item 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?.
    \label{sec:10}
    \item Colins, A., and Halverson, R. (2018).\textit{ Re-thinking education in the age of technology: The digital revolution and schooling in America}. Teachers College Press: New York.
    \label{sec:11}
    \item Cody, A. (2013). Will the data warehouse become every student and teacher's 'permanent record'? \textit{Education Week}, May, 20.
    \label{sec:12}
    \item Gibson, D. C., Webb, M., and Ifenthaler, D. (2015). Challenges of big data in educational assessment.\textit{Proceedings of the IADIS International Conference of Exploratory Learning in Digital Age}, 92-100.
    \label{sec:13}
    \item Sultan, N. (2010). Cloud computing for education: a new dawn? 30 International Journal of Information Management  109.
    \label{sec:14}
    \item Fitzgerald, B. (2014). Data collection isn't new. And it predates common core.Funny Monkey.
    \label{sec:15}
    \item Zeide, E. (2016). 19 Times Data Analysis Empowered Students and Schools: Which Students Succeed and Why?

    Project Unicorn.

\end{enumerate}
\begin{figure}
    \centering
    \includegraphics[width=15cm]{{"images/Kratos.jpg}}
    \caption{Kratos Architecture}
    \label{fig:my_label}
\end{figure}
\end{document}
